---
title: "Spotify Streaming History Analysis"
author: "@clearisawreck"
output: html_notebook
---

# Front Matter

```{r}
# clean up workspace environment
rm(list = ls())

# all packages used for this project
library(tidyverse)
library(lubridate)
library(mosaic)
library(knitr)
```

# Data Access

### Import Data Sources

```{r}
# initializing streaming history csv files for me
clare0 <- read_csv("clare0.csv")
clare1 <- read_csv("clare1.csv")
clare2 <- read_csv("clare2.csv")

# initializing streaming history csv files for my dad
dad0 <- read_csv("dad0.csv")
dad1 <- read_csv("dad1.csv")
dad2 <- read_csv("dad2.csv")

# reading spotify song database
SpotifySongs <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv')
```

### Inspecting Data Intake

```{r}
# inspecting streaming history data tables
str(clare0)
head(clare0)

# inspecting spotify songs data set
str(SpotifySongs)
head(SpotifySongs)
```

# Initial Data Wrangling

### Combine Streaming History Data Sets

```{r}
ClareHistory <- bind_rows(clare0, clare1,  clare2)
DadHistory <- bind_rows(dad0, dad1, dad2)
```

### Join Data Sets

```{r}
# use left_join to combine streaming history data sets with spotify songs data set
ClareStreamingData <- ClareHistory %>%
  left_join(SpotifySongs, by = c("trackName" = "track_name", "artistName" = "track_artist")) %>%
  distinct(endTime, trackName, .keep_all = TRUE) # remove duplicates by looking at endTime and trackName repetition

DadStreamingData <- DadHistory %>%
  left_join(SpotifySongs, by = c("trackName" = "track_name", "artistName" = "track_artist")) %>%
  distinct(endTime, trackName, .keep_all = TRUE)
```

### Examine Joined Data Sets

```{r}
str(ClareStreamingData)
head(ClareStreamingData)
```

# Examine Top Genres and Subgenres

### Data Wrangling (Genres)

```{r}
ClareGenres <- ClareStreamingData %>%
  group_by(playlist_genre) %>%
  summarize(count = n()) %>%
  na.omit() %>%
  mutate(clare = (count/sum(count))*100)

DadGenres <- DadStreamingData %>%
  group_by(playlist_genre) %>%
  summarize(count = n()) %>%
  na.omit() %>%
  mutate(dad = (count/sum(count))*100)

# combining our respective genre data sets and using pivot_longer so graphing is easier
CombinedGenres <- ClareGenres %>%
  full_join(DadGenres, by = c("playlist_genre" = "playlist_genre")) %>%
  select(playlist_genre, clare, dad) %>%
  pivot_longer(cols = c(clare, dad), names_to = "who", values_to = "percentage")
``` 

### Data Visualization (Genres)

```{r}
ggplot(CombinedGenres) + geom_col(aes(x = reorder(playlist_genre, -percentage), y = percentage, group = playlist_genre, fill = who)) + labs(x = "genre") + ggtitle("Clare and Dad Top Genres")
```

### Data Wrangling (Subgenres)

```{r}
ClareSubgenres <- ClareStreamingData %>%
  group_by(playlist_subgenre) %>%
  summarize(count = n()) %>%
  na.omit() %>%
  mutate(clare = (count/sum(count))*100)

DadSubgenres <- DadStreamingData %>%
  group_by(playlist_subgenre) %>%
  summarize(count = n()) %>%
  na.omit() %>%
  mutate(dad = (count/sum(count))*100)

CombinedSubgenres <- ClareSubgenres %>%
  full_join(DadSubgenres, by = c("playlist_subgenre" = "playlist_subgenre")) %>%
  select(playlist_subgenre, clare, dad) %>%
  pivot_longer(cols = c(clare, dad), names_to = "who", values_to = "percentage")
```

### Data Visualization (Subgenres)

```{r}
ggplot(CombinedSubgenres) + geom_col(aes(x = reorder(playlist_subgenre, -percentage), y = percentage, group = playlist_subgenre, fill = who)) + labs(x = "subgenre") + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Clare and Dad Top Subgenres")
```

# Examining Total Listening Time

### Data Wrangling (Minutes Streamed)

```{r}
ClareStreamingData %>%
  summarize(ClareTime = sum(msPlayed)/60000)

DadStreamingData %>%
  summarize(DadTime = sum(msPlayed)/60000)
```

# Examine Top Artists

### Data Wrangling (Top Artists by Number of Songs Listened To)

```{r}
ClareArtistsCount <- ClareStreamingData %>%
  group_by(artistName) %>%
  summarize(count = n()) %>% # getting artist rank by count
  arrange(desc(count))

DadArtistsCount <- DadStreamingData %>%
  group_by(artistName) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

knitr::kable(head(ClareArtistsCount), caption = "Clare Top Artists by Song Count")
knitr::kable(head(DadArtistsCount), caption = "Dad Top Artists by Song Count")
```

### Data Wrangling (Top Artists by Time Listened To)

```{r}
ClareArtistsMinutes <- ClareStreamingData %>%
  group_by(artistName) %>%
  summarize(minutes = sum(msPlayed)/60000) %>%
  arrange(desc(minutes))

DadArtistsMinutes <- DadStreamingData %>%
  group_by(artistName) %>%
  summarize(minutes = sum(msPlayed)/60000) %>%
  arrange(desc(minutes))

knitr::kable(head(ClareArtistsMinutes), caption = "Clare Top Artists by Minutes Listened")
knitr::kable(head(DadArtistsMinutes), caption = "Dad Top Artists by Minutes Listened")
```

# Examine Top Time Periods

### Data Wrangling (Isolate Years)

```{r}
ClareYears <- ClareStreamingData %>%
  mutate(date = ymd(track_album_release_date)) %>%
  na.omit() %>%
  mutate(year = year(date)) %>%
  select(year)

DadYears <- DadStreamingData %>%
  mutate(date = ymd(track_album_release_date)) %>%
  na.omit() %>%
  mutate(year = year(date)) %>%
  select(year)
```

### Data Wrangling (Group by Decades)

```{r}
ClareDecades <- ClareYears %>%
  mutate(decade = floor(year/10)*10) %>% # using floor division to get the decade in place of the year
  group_by(decade) %>%
  summarize(count = n()) %>%
  mutate(clare = count/sum(count)*100)

DadDecades <- DadYears %>%
  mutate(decade = floor(year/10)*10) %>%
  group_by(decade) %>%
  summarize(count = n()) %>%
  mutate(dad = count/sum(count)*100)
```

### Data Visualization (Decades)

```{r}
Decades <- ClareDecades %>%
  full_join(DadDecades, by = c("decade" = "decade")) %>%
  select(decade, clare, dad) %>%
  pivot_longer(cols = c(clare, dad), names_to = "who", values_to = "percentage")

ggplot(Decades) + geom_col(aes(x = decade, y = percentage, group = decade, fill = who)) + xlab("decade") + ggtitle("Clare and Dad Popular Decades")
```

# Examine Times of Day Listening

### Data Wrangling (Hours of the Day)

```{r}
ClareHour <- ClareStreamingData %>%
  mutate(date = ymd_hms(endTime), hour = hour(date)) %>% # hours are selected in a range from 0 - 23, rather than 12am to 11pm
  select(hour) %>%
  group_by(hour) %>%
  summarize(clare = n())

DadHour <- DadStreamingData %>%
  mutate(date = ymd_hms(endTime), hour = hour(date)) %>%
  select(hour) %>%
  group_by(hour) %>%
  summarize(dad = n())

Hours <- ClareHour %>%
  full_join(DadHour, by = c("hour" = "hour")) %>%
  pivot_longer(cols = c(clare, dad), names_to = "who", values_to = "count")
```

### Data Visualization (Hours of the Day)

```{r}
ggplot(Hours) + geom_line(aes(x = hour, y = count, linetype = who))
```

# Examine Song Retention

```{r}
ClareRetention <- ClareStreamingData %>%
  mutate(retention = (msPlayed/duration_ms)*100) %>%
  na.omit() %>%
  select(retention) %>%
  summarize(avg_retention = mean(retention))

DadRetention <- DadStreamingData %>%
  mutate(retention = (msPlayed/duration_ms)*100) %>%
  na.omit() %>%
  select(retention) %>%
  summarize(avg_retention = mean(retention))

head(ClareRetention)
head(DadRetention)
```

# Examine Similarity

### Data Wrangling (Similarity)

```{r}
# getting list of unique songs and their stream counts
ClareUnique <- ClareStreamingData %>%
  group_by(artistName, trackName) %>%
  select(artistName, trackName) %>%
  summarize(clarecount = n()) %>%
  arrange(desc(clarecount))
  
DadUnique <- DadStreamingData %>%
  group_by(artistName, trackName) %>%
  select(artistName, trackName) %>%
  summarize(dadcount = n()) %>%
  arrange(desc(dadcount))

knitr::kable(head(ClareUnique), caption = "Clare's Top Streamed Songs")
knitr::kable(head(DadUnique), caption = "Dad's Top Streamed Songs")
```

```{r}
# getting our total stream counts
claretotalstreams <- nrow(ClareStreamingData)
dadtotalstreams <- nrow(DadStreamingData)

# joining our data sets 
Unique <- ClareUnique %>%
  inner_join(DadUnique, by = c("artistName" = "artistName", "trackName" = "trackName")) %>% # i use inner_join so that I only keep the songs each of us have in common
  transform(incommon = pmin(clarecount, dadcount)) %>% # i assign the variable "incommon" to be the number of streams we had in common, by taking the minimum streams between each of us to be an identical stream
  summarize(claresimilarity = sum(incommon)/claretotalstreams*100, dadsimilarity = sum(incommon)/dadtotalstreams*100)

knitr::kable(head(Unique), caption = "Similarity Scores")
```



